### 2014-09-11 paper/paper.Rmd
hyphen
section 3

Although not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each gender only. We would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score. There is far more individual variability than the overall average differences: who you are is more important than your demographic.

@@

For each country we have computed the average math test score by age that the children started school. In additon, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.

Within a country, looking at the average math score for each start age, the score typically declines, especially after age 6. This is almost uniformly the pattern across all of the countries in the study. For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than another other starting age group. School start age matters, that children need to start school in the 4-6 age range to expect better math scores, on average.

@@

# 3. Teachers, computers, and libraries at school make a difference in math scores

Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction. The plot below examines at three of these: lack of teachers, lack of computers and shortage of library materials. These factors are compared with math, reading and science scores from the student table, across countries. For most countries, scores decrease on average the greater the shortage.

@@

```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
qplot(levels, math, data = df.m, facets = ~name, 
      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
      theme(legend.position="bottom") 
```

>>>>

Although not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each gender only. We would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys' score. There is far more individual variability than the overall average differences - who you are is more important than your demographic.

@@

For each country, we have computed the average math score by age that the children started school. In addition, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia, and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, e.g. in Japan all children start at age 6.

Within a country, looking at the average math score for each start age, the score typically declines, especially after age 6. This is almost uniformly the pattern across all of the countries in the study. For several countries - Russia, Denmark, Bulgaria - if a child starts school at age 4 their math scores were much better on average than any other age group. School start age matters. Children need to start school in the 4-6 age range to expect better math scores, on average.

@@

# 3. Shortages in teachers, computers, and libraries at school make a difference in math scores in a few countries

Within the PISA dataset, the school table has information about shortages on items viewed as being important for good instruction. The plot below examines three of these: lack of teachers, lack of computers, and shortage of library materials. The barcharts show the overall counts of shortages in schools. Teacher shortage is less comman than material shortages.

These factors are compared with math scores from the student table, across countries. In most countries, scores don't change mcuh as shortages get worse. Scores decrease by 10-20 points onaverage in countries like Australia, China, Uruguay, Turkey, UAE, USA. Teacher shortages appear to have impact in a few countries like Greece, Slovakia, and Serbia.

@@

```{r shortagesbars, fig.width=10, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
df$SC14Q02 <- as.character(df$SC14Q02)
df$SC14Q07 <- as.character(df$SC14Q07)
df$SC14Q10 <- as.character(df$SC14Q10)
df$SC14Q02[df$SC14Q02=="A lot"] <- "To some extent"
df$SC14Q07[df$SC14Q07=="A lot"] <- "To some extent"
df$SC14Q10[df$SC14Q10=="A lot"] <- "To some extent"
df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2))
df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2))
df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2))
df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
#      ylab="Math Score", geom=c("point","line"), group=name)
df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
#      ylab="Math Score", geom=c("point","line"), group=name)
df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
colnames(df2)[2] <- "levels"
df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
colnames(df3)[2] <- "levels"
df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
colnames(df4)[2] <- "levels"
df.all <- merge(df2, df3)
df.all <- merge(df.all, df4)
df.m <- melt(df.all, id=c("name", "levels"))
colnames(df.m)[4] <- "math"
dqplot(levels, math, data = df.m, facets = ~name, 
      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
      theme(legend.position="bottom") 
```
